---
title: "Yield Tables in ForestElementsR"
author: "Peter Biber"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Yield Tables in ForestElementsR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1 Introduction
Yield tables might be the most important tools forest science has ever provided
to forest practice. They represent an invaluable wealth of empirical data that
has been condensed into tabular descriptions of forest stand growth that are 
easy to work with. Since about three decades, however, they have been 
increasingly frowned upon for several reasons, among them their limited 
applicability to mixed species stands, their inherent restriction to even aged 
stands only, and  their representation of forest growth under environmental 
conditions that have meanwhile changed, and cannot be considered as stable as 
they were at the time when the data behind the tables were collected. 
Nevertheless, when handled wisely, yield tables can still be highly valuable 
assets in our tool kit. Therefore, we developed a generic implementation of 
yield tables in *ForestElementsR* that is designed to work equally well with 
yield tables of different origin with different specifications and descriptive 
variables. A collection of yield tables for important tree species in Central 
Europe has already been implemented and comes readily with *ForestElements*; 
this collection will certainly keep growing. All readily available tables are 
listed in the documentation, you can look them up with ```?yield_tables```. 
The purpose of this vignette is to explain how to work with our yield table 
implementation.

```{r}
# Before we start, let's attach the package
library(ForestElementsR)
```


## 2 Yield Table Representation in ForestElementsR
Let's have a look to a classic yield table, Wiedemann's 1943 table for Scots
pine (*Pinus sylvestris*). The bulk information used for our implementation was 
the version published by the Bavarian State Forest Administration 
[@hilfstafeln_bavaria_2018], some information that was missing there, was taken
from the version in Schober's yield table collection 
[@ertragstafeln_schober_1975] in addition. As any yield table in 
ForestElementsR, the Wiedemann table for Scots pine is an S3 object of class
```fe_yield_table```. its representation in the package is called
```fe_ytable_pine_wiedemann_moderate_1943```. 

```{r}
class(fe_ytable_pine_wiedemann_moderate_1943)
```


Such an object is essentially a 
list containing some metadata followed by the actual data. A basic overview can
be obtained with:

```{r}
fe_ytable_pine_wiedemann_moderate_1943 |> summary()
```

The first six list elements represent the metadata, the seventh element, 
```values```, comprises the actual data.


### 2.1 Metadata
Let's have a look at the metadata first:

```{r}
fe_ytable_pine_wiedemann_moderate_1943[1:6] # Elements 1-6 contain metadata
```

The list element ```name_orig``` contains the established original name of the 
yield table in the language it was originally published in. The second item, 
```name_international``` is an English translation of the original name. The 
item ```site_indexes``` is a numeric vector of all site index values that are 
actually represented in the table with values. The Wiedemann pine table is using
the traditional European approach of *relative site indexing*, i.e. site index 
1.0 stands for the best site conditions, while 4.0 or greater represent 
considerably weak sites. A site index of 1.5 would be in the middle between the 
best and the second best site represented in the table. In its original form, 
this traditional site index numbering actually uses roman numerals before and 
arabic ones after the decimal point (e.g. II.3 = 2.3), but implementing this 
simply was not worth the effort. More important, our implementation works 
equally with yield tables that follow the approach of absolute site indexing 
(which is preferable from our point of view, but not frequently found in German 
yield tables). Have a look at the ```site_indexes``` entry of the yield table
by Ernst Assmann and Friedrich Franz for Norway spruce (as taken from
@hilfstafeln_bavaria_2018):

```{r}
fe_ytable_spruce_assmann_franz_mean_yield_level_1963$site_indexes
```

These site indexes represent a stand's dominant height in meters at an age of
100 years. So, a stand of site index 40 after Assmann and Franz would be 
expected to be 40 m high when it is 100 years old. As you will see below, we
have implemented the option to express site quality in terms of such an absolute
site index also for yield tables with a relative site index system. Back to the
Wiedemann pine table, the next element of the metadata is called 
```site_index_variable```. This is a character vector with - typically - one 
element that provides the name of the variable which is used to determine the 
site index. As for most German yield tables, for the Wiedemann table it is the 
quadratic mean height in meters, abbreviated as "h_q_m". This means when the
quadratic mean height (and the age) of a Scots pine stand is known, its site 
index can be determined with this yield table. Again, the Assmann-Franz table
can serve as an example for other possibilities:

```{r}
fe_ytable_spruce_assmann_franz_mean_yield_level_1963$site_index_variable
```

This table was designed for site indexing based on the stands dominant height,
here defined as the quadratic mean height of the 100 tallest trees per hectare,
called "h_100_m" in our implementation. But in addition, this table also 
contains the usual quadratic mean stand height, "h_q_m", so we allow both 
variables to be used for finding a stand's site index. The next metadata element
we called ```mai_variable```. This relates to the concept of expressing site 
indexes in terms of the mean annual increment (the German term for this is 
"DGZ-Bonitierung"). The idea of that approach is to use a yield table with the
classic input (stand height, age), but to express the site quality in terms of
the expected mean annual increment ("m.a.i.", German abbreviation "dGz") at a 
given age or at its maximum (see more details below). The last metadata item,
```age_coverage``` is a numeric vector of all ages for which the table actually
provides data. Note, that this is an "outer" age coverage, i.e. not all site
indexes given in the table must necessarily cover the whole range of values.
Often, poor site indexes start in the table at a higher age than better ones.
Sometimes, the age interval might be different for different site indexes. E.g.
the Wiedemann pine table usually provides values in five year age intervals, but
in ten-year intervals only for site indexes 5.5 and 6.0. This is also not a
problem with the age coverage to be given in the metadata; it always relates
to the highest temporal resolution available in the table.

### 2.2 Actual Yield Table Data
The yield table data themselves are stored in the list element called 
```values```. This itself is a named list of matrices, one matrix for each 
variable given in the yield table, the list element names being the names of the
variables. Note, that ```fe_yield_table``` objects are not required to have all
an identical set of variables. In order to find out which variables are given in
a yield table, you could obtain their names as follows:

```{r}
fe_ytable_pine_wiedemann_moderate_1943$values |> names()
```

The order of the variables is not related to their importance, e.g. the first
variable listed here, *h_q_m_si_pluis_025*, is the lower threshold height for 
a given site index as by default provided in the Bavarian state forest yield 
table editions (@hilfstafeln_bavaria_2018, @hilfstafeln_bavaria_1990). For 
electronic site indexing this variable is not required, we kept it, however,
test and documentation purposes. Of central importance, in contrast, are the
variables which are used for the actual site indexing. In the Wiedemann pine
table, this is  ```h_q_m"```, the quadratic mean stand height, as indicated 
in the metadata. As for any other variable in the yield table, it can be simply 
displayed as follows:

```{r}
fe_ytable_pine_wiedemann_moderate_1943$values$h_q_m
```

The columns of the matrix correspond to the site indexes given in the metadata.
They are represented in the column names with the prefix "si_". Each matrix row
stands for a specific stand age, whereby the stand ages are the row names. The 
NA values are places where no value is given in the table, typically with low 
site indexes at younger ages or with broader age intervals. All other variables
in the table are given as a matrices of exactly the same structure, take e.g.
the periodic annual increment ```pai_m3_ha_yr```:

```{r}
fe_ytable_pine_wiedemann_moderate_1943$values$pai_m3_ha_yr
```

Before we come to visualizing yield table contents, let us shortly explain the
variable names used in the Wiedemann pine table, as we are sticking as close as
possible to this naming concept in any yield table we make a part of 
ForestElementsR. Note, however, that not all of these variables are necessarily
contained in all tables we have included so far; some tables might also contain
additional variables and (unfortunately) use other than metric units. For 
creating a working ```fe_yield_table``` object it is not necessary to stick to a
certain naming concept although we highly recommend to do so. Let's explain the 
variable names given above, beginning whith those that are common in most yield 
tables:

* **h_q_m**: Quadratic mean stand height (m)
* **d_q_cm**: Quadratic mean diameter at breast height (cm)
* **n_ha**: Number of trees per ha
* **ba_m2_ha**: Stand basal area (m²/ha)
* **v_m3_ha** : Stand wood volume (m³/ha)
* **tvp_m³_ha**: Total volume production (m³/ha)
* **mai_m³_ha_yr**: Mean annual volume increment (m³/ha/yr), i.e. the total 
  volume production divided by the stand age
* **pai_m3_ha_yr**: Periodic mean annual increment (m³/ha/yr), i.e. the 
  difference of the total volume production for two subsequent points in time
  divided by the corresponding time interval 
* **pai_perc_yr**: Annual relative periodic increment (%/yr)  

As the Wiedemann pine table was imported from the edition provided by the
Bavarian State Forest Service [@hilfstafeln_bavaria_2018], there are some
peculiarities: Some of the variable names carry the prefix "red_". This means,
while volume-related values are usually, to be understood as m³ standing wood 
over bark, the prefix "red_" ("reduced") indicates, that such wood volumes are 
to be understood under bark and after subtracting losses due to the harvest.
This explains the variable names **red_v_m3_ha**, **red_mai_m3_ha_yr**, and
**red_pai_m3_ha_yr**. Another peculiarity of the Bavarian table edition is the
scarce information about the removal stand. For a given stand age, it provides 
only the harvest volume for the subsequent decade, also under bark, and after
harvest, hence the name **red_pre_yield_m3_ha_10yr**. More information about the
removal stand in yield tables will be provided (if available from other table
editions such as @ertragstafeln_schober_1975) in future versions of 
*ForestElementsR*.

## 3 Visualizing Yield Table Contents
For visualization purposes whe have written a plot method for 
```fe_yield_table``` objects.

```{r, fig.width=4.9, fig.height=4.9, fig.align="center"}
# Plot the quadratic mean height (site index fan)
fe_ytable_pine_wiedemann_moderate_1943 |> plot()
```

When the plot function is called without further arguments, it displays the 
first variable listed as ```site_index_variable``` in the object's metadata.
Typically this results in a display of a "site index fan". All other variables 
contained in the ```fe_yield_table``` object can be plotted by specifiying 
their name in the call:

```{r, fig.width=4.9, fig.height=4.9, fig.align="center"}
# Plot the periodic annual increment
fe_ytable_pine_wiedemann_moderate_1943 |> plot(variable = "pai_m3_ha_yr")
```

The plots are *ggplot* objects whose appearance can be partly modified in a 
post-hoc manner if required, e.g.

```{r, fig.width=4.9, fig.height=4.9, fig.align="center"}
# Plot the tree number per ha and modify the plot's appearance
# 1 catch the plot
v_plot <- fe_ytable_pine_wiedemann_moderate_1943 |> plot(variable = "n_ha")

# 2 make adjustments as allowed by ggplot2, e.g.
v_plot + 
  ggplot2::theme_classic() +     # Appearance similar to R base graphics
  ggplot2::scale_y_log10() +     # Logarithmic scale for vertical axis
  ggplot2::ylab("Tree Number per hectare") # More explaining axis label
```


## 4 Practical Work With Yield Tables in ForestElementsR
### 4.1 Site Indexing
The fundamental task in working with yield tables is finding the site index of a 
given stand. This requires two input variables, the stand's age, and the stand's
height, defined as (one of) the yield table's site index variable. For the 
Wiedemann pine table this is the quadratic mean height in meters. The function
to be used is called ```site_index()```. Let's have a look at some examples:

```{r}
ytab <- fe_ytable_pine_wiedemann_moderate_1943 # store the yield table in a
                                               # variable with a shorter name
                                               # for convenience

# Assume a stand age of 73 years and a quadratic mean height of 19.7 m
si <- site_index(age = 72, size = 19.7, ytable = ytab, si_variable = "h_q_m")
si # The stand's relative site index according to the table
si |> round(digits = 1) # usually relative site indexes are rounded to one digit

# Same height, but twenty years younger
site_index(age = 52, size = 19.7, ytable = ytab, si_variable = "h_q_m") |>
  round(digits = 1)

# Same height, but thirty years older
site_index(age = 102, size = 19.7, ytable = ytab, si_variable = "h_q_m") |>
  round(digits = 1)
```

Input values for stand ages that are not exactly given in the table, as in the 
examples above, are made use of by linear interpolation between the values given
in the corresponding matrix. Site index values that are outside the site index
fan, are obtained by linear extrapolation. This is the case in the second 
example, where we obtain a site index of 0.9 which reflects better site 
conditions than assumed by the best site index given in the table (i.e. 1.0).
If a stand age provided by the user is beyond the age coverage of the yield
table (as provided in the ```fe_yield_table``` object's slot 
```age_coverage```), a warning is issued, and the age that is internally used is
set to the nearest value covered in the table:

```{r}
site_index(age = 171, size = 30.4, ytable = ytab, si_variable = "h_q_m")
site_index(age = 12, size = 9.2, ytable = ytab, si_variable = "h_q_m")
```

Obviously, site indexing can be done on input vectors, in the following example
we are doing this by way of the function ```map2_dbl``` from the *purrr*
package:

```{r}
age <- c(35,   25,   120, 75,   42,    53)
h_q <- c(14.3, 9.1, 22.2, 13.6, 11.0,  21.7)

si_vec <- purrr::map2_dbl(
  .x = age, 
  .y = h_q, 
  .f = function(x, y, yt, si_v) site_index(x, y, yt, si_v), 
  yt = ytab, 
  si_v = "h_q_m"
) |>
  round(digits = 1)

si_vec
```

In order to convert this site index into a mean annual increment (mai) site 
index (German: "DGZ-Bonität"), there are two options. First, the conversion can 
be made to the expected mai at a given age, typically 100 years. This is 
achieved with the function ```si_to_mai_age```:

```{r}
# Obtain the expected mean annual increment at age 100 for a stand with the
# relative site index 1.2
si_to_mai_age(
  si = 1.2, mai_variable = "mai_m3_ha_yr", age = 100, ytable = ytab
)

# Do the same for the site index vector we generated above
si_vec

purrr::map_dbl(
  .x = si_vec,
  .f = function(x, ma_v, age, yt) si_to_mai_age(x, ma_v, age, yt),
  ma_v = "mai_m3_ha_yr",
  age  = 100,
  yt   = ytab
)
```

The second option is not to determine the mai site index at a given age, but to
take the maximum mai, that does typically occur not at the same age among 
different site qualities (earlier on good sites, later on poor sites).
The function ```si_to_mai_max``` was written for that purpose:

```{r}
# Obtain the expected maximum mean annual increment for a stand with the
# relative site index 1.2
si_to_mai_max(
  si = 1.2, mai_variable = "mai_m3_ha_yr", ytable = ytab
)

# Do the same for the site index vector we generated above
si_vec

purrr::map_dbl(
  .x = si_vec,
  .f = function(x, ma_v, yt) si_to_mai_max(x, ma_v, yt),
  ma_v = "mai_m3_ha_yr",
  yt   = ytab
)
```

Note that in some cases, especially when extrapolating with site indexes that 
are above the best site covered in a yield table, the obtained mai max site 
index might be smaller than the mai site index obtained for an age that is near
the maximum mai. Unfortunately, such subtle inconsistencies are in the very 
nature of yield tables in general, and cannot be avoided. In order to convert 
relative site indexes into absolute ones, e.g. the expected stand height at an
age of 100 years, you can use the function ```ytable_lookup``` which is the 
general means to draw values out of yield tables, once the site index is known:

```{r}
# Obtain the expected stand height at age 100 for a stand with the relative
# site index 1.2 (this actually converts a relative into an absolute site index)
si_abs <- ytable_lookup(age = 100, si = 1.2, variable = "h_q_m", ytable = ytab)
si_abs
si_abs |> round(digits = 1)

# Let's to this again for the whole site index vector from above.
# Reasonably, we obtain warnings, if we use site indexes beyond range of the 
# yield table
si_abs_vec <- purrr::map_dbl(
  .x = si_vec,
  .f = function(x, age, var, yt) ytable_lookup(age, x, var, yt),
  age = 100,
  var = "h_q_m",
  yt   = ytab
)

si_abs_vec |> round(digits = 1)

```

### 4.2 Extracting Values From Yield Tables
The key function for extracting desired values from yield tables, once the site
index is known, is ```ytable_lookup``` as already shown above when the task was
to obtain absolute site indexes from relative ones. Let us give some more 
examples here:

```{r}
# Get the periodic annual increment:
ytable_lookup(age = 100, si = 1.2, variable = "pai_m3_ha_yr", ytable = ytab)

# Try to go beyond the table's age coverage (raises warning)
ytable_lookup(age = 170, si = 1.2, variable = "pai_m3_ha_yr", ytable = ytab)

# ... standing volume
ytable_lookup(age = 73, si = 2.4, variable = "v_m3_ha", ytable = ytab)

# Use a site index above the table's coverage (raises warning)
ytable_lookup(age = 73, si = 0.4, variable = "v_m3_ha", ytable = ytab)

# ... basal area
ytable_lookup(age = 41, si = 3.4, variable = "ba_m2_ha", ytable = ytab)

# Use a site index below the table's coverage
ytable_lookup(age = 41, si = 6.2, variable = "ba_m2_ha", ytable = ytab)
```

This way, any variable that is contained in the ```fe_yield_table``` object can
be easily accessed. Vectorized application is available with standard R 
functionality (we prefer the functions from the *purrr* package, but the *apply*
family of standard R works as well).


## 5 Importing Yield Table Data and Make them fe_yield_table Objects
The key function for transoforming your own data into valid and fully functional 
```fe_yield_table``` objects is ```fe_yield_table```. We will add some more 
explanations here; for the time being, be referred to the function's 
documentation. You can access it with ```?fe_yield_table```.

## References
